---
title: "Image Level Analysis"
author: "Cody Heiser"
date: "7/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(GPs=function(before, options, envir){
  if (before){
    cex=1.5
    par(mgp=c(1.7,.7,0), lwd=1.5, lend=2,
        cex.lab=0.8*cex, cex.axis=0.8*cex, cex.main=1*cex,
        mar=c(2.8,1.8,1.8,0), bty='l', oma=c(0,0,2,0))}
})
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 4, GPs=TRUE, cache=TRUE, cache.lazy = FALSE)

source("~/git/atlasAnalysis/tissue_clustering_utils.r")
library(docstring)
library(parallel)
library(class)
library(fda)
library(lme4)
library(lmerTest)
library(emmeans)
library(sjPlot)
```

## 1. Setup

This first code chunk sets up the data frame `sr` that contains file paths to all of the images.
We need to get a mask from Joe that excludes regions where the tissue came off the slide across the rounds of staining.

```{r dataSetup, echo=FALSE}
wd <- "~/biostat_clustering/"  # working base directory

# markers to use:
# These are modifiable. location markers are used for clustering
locationMarkers <- c(
  #'ACTININ',
  #'BCATENIN',
  #'CGA',
  'COLLAGEN',
  #'COX2',
  'DAPI',
  #'ERBB2',
  #'GACTIN',
  #'HLAA',
  #'LYSOZYME',
  'MUC2',
  'NAKATPASE',
  'OLFM4',
  'PANCK',
  #'PCNA',
  #'PDL1',
  #'PEGFR',
  #'PSTAT3',
  #'SMA',
  #'SNA',
  'SOX9',
  'VIMENTIN'
  )

# immune markers for downstream tissue infiltration analysis
immuneMarkers <- c(
  #'FOXP3',
  #'CD3D',
  #'CD4',
  #'CD8',
  #'CD11B',
  #'CD20',
  #'CD45',
  #'CD45B',
  #'CD68'
  )

# define slide names in directory structure
slides <- c(
  "MAP00083_0000_02_01",
  "MAP00347_0000_03_02"
  #"MAP00377_0000_01_01",
  #"MAP00392_0000_01F_01",
  #"MAP00411_0000_01A_0",
  #"MAP00546_0000_02_02",
  #"MAP02112_0000_02_02",
  #"MAP05994_0000_01A_0",
  #"MAP06025_0000_02_01"
  )

# use data setup function to create sr
sr <- setup_data(
  wd = wd,
  slides = slides,
  markers = locationMarkers
)

# # append tumor subtype information (optional)
# tts = read.csv(file.path(wd, 'TissueSubTypes Colon Map.csv'))
# names(tts)[1] = 'TissueID'  # match capitalization
# sr$TissueID = sr$Slide
# sr = merge(sr, tts, all.x=TRUE)
# sr$adenSSLcrc = ifelse(sr$tissueSubType %in% c('Tub', 'TV'), 'adenoma', 
#                         ifelse(sr$tissueSubType %in% c('HP', 'HP/SSL', 'SSL', 'Tub/SSL'), 'SSL', 
#                                ifelse(sr$tissueSubType %in% c('CRC'), 'CRC', NA )) )
# ss = 1:nrow(sr)  # get counter to use in loops
# # checks that mask image matches dimensions of DAPI image (some didn't)
# dimsMatch = apply(sr[ss,], 1, function(x) all(dim(raster(x['mask'])) == dim(raster(x['DAPI']))) )
# ss = ss[dimsMatch]
```

## 2. Downsampling

Here, we downsample the images as input to Kmeans clustering to define general tissue level clusters.
The downsampling returns the mean value of the pixels in the lower resolution image.
This substantially changes the distribution of the image intensities.
The `fun` argument of the downsample function can be used to change how smaller pixel values are combined into the larger value.
Median would be another reasonable choice, probably.

The code chunk below also includes initial parameters for the kmeans and smoothing.

```{r downsample}
# use downsampling function to loop through sr
sr2 <- downsample_data(
  sr,
  fact = 8,
  markers = locationMarkers,
  masks = c("mask"),#,"epiMask","strMask","tumorMask"),
  njobs = 4,
  overwrite=F
)
```

## 3. Spatial Smoothing

Here we use the gaussSmooth function to apply smoothing. It is not actually Gaussian smoothing, because the `raster` package did not have an efficient way to deal with edge effects using Gaussian smoothing, so it's actually homogeneous smoothing on a disk of radius `sigma`.
It's applied to all of the location markers, which are being used to define broad tissue classes.

The pipeline is:

1. Downsample
2. Normalize (done here before smoothing)
3. Smooth

```{r smoothing, eval=TRUE}
# use smoothing function to loop through sr
sr3 <- smooth_data(
  sr2,
  sigma=50,
  markers=c("COLLAGEN","DAPI","MUC2","NAKATPASE","OLFM4","PANCK","SOX9","VIMENTIN"),
  njobs=4,
  overwrite=F
)
```

## 4. Spatial Clustering

* We subsample at a factor of ``r fact``, so we sample 1/``r fact^2`` of the data on a grid.
* We used mean division with an offset of ``r offset`` and a `log10` transformation.
* We also used mean division without a `log10` transformation too.

This first chunk runs the k-means on the log-transformed data.
It then uses that model to assign the labels to all of the pixels and writes them out as the `locationLabel` images.

```{r kmeans, eval=TRUE}
# image(raster(sr$MUC2[1]))
kMeansMarkers <- paste(locationMarkers, 'smooth', sep="_")

sr4 <- cluster_data(
  sr3,
  k=4,
  subsamp=8,
  transform=log10,
  offset=0.5,
  markers=kMeansMarkers,
  seed=NA,
  out.prefix='locationLabel'
)
```

This second chunk runs k-means on the untransformed data and then assigns the values to the `locationLabelUntransformed` images.

```{r kmeansUntransformed, eval=FALSE}
sr5 <- cluster_data(
  sr4,
  k=4,
  subsamp=8,
  transform=function(x) x,
  offset=0,
  markers=kMeansMarkers,
  seed=NA,
  out.prefix='locationLabelUntransformed'
)
```
